---
title: "Arrow output"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`dplyr::collect()` turns a query result into an R data frame. Sometimes you
want the result in [Apache Arrow](https://arrow.apache.org/) format instead:

- to keep working on it with the arrow package,
- to save it as Parquet files on your own disk,
- to process a result that is too large to hold in memory,
- or to hand it to another tool that reads Arrow data.

quak has two functions for this. `collect_arrow()` returns the whole result
at once. `stream_arrow()` returns it in batches.

## Setup

Both functions need the nanoarrow package. quak offers to install it the
first time you call one of them. The arrow package is optional, but most of
the examples below use it.

```{r}
install.packages(c("nanoarrow", "arrow"))
```

The examples use a Delta table on Azure:

```{r}
library(quak)

conn <- az_conn()
az_set_chain_secret(conn, chain = "cli")

sales <- tbl_delta(conn, "abfss://container@account/path/sales")
```

## Choose a function

| You want | Use | You get |
|---|---|---|
| An R data frame | `dplyr::collect()` | A tibble |
| Arrow data, all in memory | `collect_arrow()` | A `nanoarrow_array_stream`, already read |
| Arrow data, one batch at a time | `stream_arrow()` | A `nanoarrow_array_stream` |

All three run the query in DuckDB. Filter, select and summarise before you
collect, so that less data leaves DuckDB:

```{r}
large_sales <- sales |>
  dplyr::filter(amount > 100) |>
  dplyr::select(id, region, amount)
```

Both Arrow functions check the connection before they run, the same way
`collect()` does. They fail with a clear error if the connection is closed
or the `azure` extension is not loaded.

## Collect the whole result

`collect_arrow()` runs the query and reads the whole result into memory. It
returns the rows as an Arrow stream whose batches are already loaded. The
query has finished by the time it returns, so the connection is free for
other queries.

```{r}
arr <- collect_arrow(large_sales)
#> ℹ Collecting data from Azure...
#> ✔ Done. 82500 rows collected in 93 ms.
```

Like every Arrow stream, the result can be read only once. Convert it
straight away and keep the converted object.

For an R data frame, use `as.data.frame()` or `tibble::as_tibble()`. Neither
needs the arrow package:

```{r}
df <- as.data.frame(arr)
head(df, 3)
#>   id region amount
#> 1 67  south  100.5
#> 2 68   east  102.0
#> 3 69  north  103.5
```

The column types mostly match what `collect()` returns. `INTERVAL` columns
are the main exception. nanoarrow cannot convert them, so cast them in the
query or convert through arrow. See `?collect_arrow` for the full list.

### Continue with arrow

To keep working in Arrow instead, convert the result with
`arrow::as_arrow_table()`. Unlike the stream, a Table can be used as often
as you like. You can use arrow's own dplyr support on it, which runs in
Arrow's query engine rather than in R:

```{r}
tab <- arrow::as_arrow_table(collect_arrow(large_sales))

tab |>
  dplyr::group_by(region) |>
  dplyr::summarise(total = sum(amount)) |>
  dplyr::collect()
#> # A tibble: 3 × 2
#>   region    total
#>   <chr>     <dbl>
#> 1 south  3423734.
#> 2 east   3423750
#> 3 north  3423766.
```

Or save it as a local Parquet file:

```{r}
arrow::write_parquet(tab, "sales_large.parquet")
```

## Stream the result

`stream_arrow()` runs the query and returns a stream. The stream produces
batches of at most `chunk_size` rows, one at a time, as you read them. Only
the current batch is in memory, so the whole result never has to fit.

```{r}
stream <- stream_arrow(sales, chunk_size = 100000)
```

As with `collect_arrow()`, a stream can be read only once. Once a batch has
been read, it cannot be read from the stream again.

### Hand the stream to arrow

`arrow::as_record_batch_reader()` turns the stream into an arrow
`RecordBatchReader`. Most arrow functions accept one, and they read it
batch by batch.

Copy a large table from Azure to a partitioned Parquet dataset on your own
disk, without holding it in memory:

```{r}
reader <- arrow::as_record_batch_reader(stream_arrow(sales))

arrow::write_dataset(
  reader,
  "data/sales",
  format = "parquet",
  partitioning = "region"
)

list.files("data/sales", recursive = TRUE)
#> [1] "region=east/part-0.parquet"  "region=north/part-0.parquet"
#> [3] "region=south/part-0.parquet"
```

Then open the local copy with `arrow::open_dataset()` and query it without
going back to Azure.

You can also run dplyr verbs on the reader. arrow runs them on each batch as
it arrives:

```{r}
reader <- arrow::as_record_batch_reader(stream_arrow(sales))

reader |>
  dplyr::filter(amount > 140) |>
  dplyr::summarise(n = dplyr::n()) |>
  dplyr::collect()
#> # A tibble: 1 × 1
#>       n
#>   <int>
#> 1 15000
```

When the same filter can run in DuckDB, apply it to `sales` before
streaming instead. DuckDB then reads less data from Azure.

### Process batches in R

To work through the batches yourself, call `stream$get_next()`. It returns
the next batch as a `nanoarrow_array`, or `NULL` when the stream is
finished:

```{r}
stream <- stream_arrow(sales, chunk_size = 100000)

total <- 0
while (!is.null(batch <- stream$get_next())) {
  df <- as.data.frame(batch)
  total <- total + sum(df$amount)
}
```

### Stop reading early

If you stop before the end, call `stream$release()`. It frees the query and
its memory straight away, instead of whenever R next collects garbage:

```{r}
stream <- stream_arrow(sales)
first <- stream$get_next()
stream$release()
```

### Keep the connection free while a stream is open

Read a stream to the end, or release it, before you run any other query on
the same connection. DuckDB ends an open stream as soon as its connection
runs another query. The stream then acts as if it had finished normally,
without an error, so the rows it had not yet returned are lost:

```{r}
stream <- stream_arrow(sales, chunk_size = 10)
stream$get_next()$length
#> [1] 10

DBI::dbGetQuery(conn, "SELECT 42")  # another query on the same connection

stream$get_next()
#> NULL
```

This also applies to a `RecordBatchReader` made from a stream, because arrow
reads it lazily. Finish with the reader, for example by letting
`arrow::write_dataset()` or `dplyr::collect()` complete, before you use the
connection again. The behaviour is reported upstream as
[duckdb/duckdb-r#2772](https://github.com/duckdb/duckdb-r/issues/2772).

If you need to query while a stream is open, use a second connection.
Queries on another connection do not affect the stream. `az_conn()` opens a
new, empty database, so register a secret on it again. Tables and views
that you created with `load_delta()` and similar functions on the first
connection are not visible there.

### Choose a batch size

`chunk_size` is the largest number of rows in one batch. The default is one
million. Smaller batches use less memory each. Larger batches mean fewer
steps and less overhead per row. Lower it when rows are wide, for example
when they hold long strings or nested columns.

DuckDB allocates each batch for `chunk_size` rows before filling it, even
when the result has fewer rows. Keep `chunk_size` modest: a value in the
hundreds of millions can reserve gigabytes of memory.

## Related functions

`arrow::to_arrow()` also turns a DuckDB lazy table into a
`RecordBatchReader`. The quak functions add the connection checks and
messages that `collect()` has, and they need only nanoarrow, not arrow.
